---
title: "Example-Extended Conditional Density Estimation Network (CDEN) Model"
author: "B. R. P. M. Banayake"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Libraries
```{r}
# Load necessary libraries
library(CaDENCE)
library(openxlsx) # For writing results to Excel
```

## Load and Explore Data
```{r}
# Load the dataset
data(FraserSediment)

# Display the first 20 rows of the dataset
head(FraserSediment, 20)

# Display the structure of the dataset
str(FraserSediment)
```

## Define Training and Testing Data
```{r}
# Extract training data (1970-1976)
x_train <- FraserSediment$x.1970.1976
y_train <- FraserSediment$y.1970.1976

# Extract testing data (1977-1979)
x_test <- FraserSediment$x.1977.1979
y_test <- FraserSediment$y.1977.1979
```

## Define Log-Normal Distribution
```{r}
lnorm.distribution <- list(
  density.fcn = dlnorm,
  parameters = c("meanlog", "sdlog"),
  output.fcns = c(identity, exp)
)
```

## Define Activation Functions
```{r}
hidden.fcn2 <- list(
  function(h) { tanh(h) },  # Activation function for first hidden layer
  function(h) { tanh(h) }   # Activation function for second hidden layer
)
```

## Train Models with Multiple Random Seeds
```{r}
# Initialize a list to store models
models <- list()

# Define seed values for reproducibility
seed_values <- 123:152  # Range of seed values

# Train models for different seeds
for (i in seed_values) {
  set.seed(i)  # Set the seed
  
  # Train the CDEN model - with two hidden layers of 2 and 3 hidden neurons 
  model <- cadence.fit_lnorm(
    x = x_train, y = y_train, iter.max = 100, hidden.neurons = c(2,3), 
    hidden.fcn = hidden.fcn2, distribution = lnorm.distribution, 
    sd.norm = 0.1, init.range = c(-0.1, 0.1), method = "Rprop", 
    n.trials = 100, trace = 1, maxit.Nelder = 2000, trace.Nelder = 0,
    swarm.size = NULL, vectorize = TRUE, delta.0 = 0.1, delta.min = 1e-06, 
    delta.max = 50, epsilon = 1e-08, range.mult = 2, step.tol = 1e-08, 
    f.target = -Inf, f.cost = cadence.cost_lnorm, max.exceptions = 500
  )
  
  # Store the model in the list
  models[[paste0("model_seed_", i)]] <- model
  cat("Model with seed", i, "completed and saved.\n")
}
```

## Extract Negative Log-Likelihood (NLL) Values
```{r}
# Initialize a data frame to store NLL values
nll_results_df <- data.frame(Seed = integer(), NLL = numeric())

# Retrieve NLL values for each model
for (i in seed_values) {
  nll_value <- attr(models[[paste0("model_seed_", i)]]$fit, "NLL")
  nll_results_df <- rbind(nll_results_df, data.frame(Seed = i, NLL = nll_value))
}

# Save results to an Excel file
write.xlsx(nll_results_df, file = "Model_NLL_Results.xlsx")
```

## Generate Predictions for Each Model
```{r}
b_values <- list()
pred_b_values <- list()

# Generate predictions using trained models
for (seed in seed_values) {
  b <- cadence.predict_lnorm(x_test, fit=models[[paste0("model_seed_", seed)]], 
                             hidden.neurons=c(2,3), hidden.fcn=hidden.fcn2)
  b_values[[paste0("b_seed_", seed)]] <- b
  pred_b <- exp(b[,1] + (((b[,2])^2)/2))
  pred_b_values[[paste0("pred_b_seed_", seed)]] <- pred_b
}
```

## Compute Error Metrics
```{r}
# Initialize a data frame to store error metrics
error_metrics_df <- data.frame(Seed = integer(), MAE = numeric(), MAPE = numeric(), MSE = numeric(), RMSE = numeric())

# Compute error metrics for each model
for (i in seed_values) {
  pred_b <- pred_b_values[[paste0("pred_b_seed_", i)]]
  mae_value <- MAE(y_test, pred_b)
  mape_value <- MAPE(y_test, pred_b)
  mse_value <- MSE(y_test, pred_b)
  rmse_value <- RMSE(y_test, pred_b)
  error_metrics_df <- rbind(error_metrics_df, data.frame(Seed = i, MAE = mae_value, MAPE = mape_value, MSE = mse_value, RMSE = rmse_value))
}

# Save error metrics to an Excel file
write.xlsx(error_metrics_df, file = "Error_Metrics_Results.xlsx")
```

## Display Sample Predictions
```{r}
# Display first 20 predicted values
head(pred_b_values$pred_b_seed_123, 20)
head(y_test, 20)

# Display last 20 predicted values
tail(pred_b_values$pred_b_seed_123, 20)
tail(y_test, 20)
```

## Summary
This R Markdown file provides a step-by-step implementation of a Conditional Density Estimation Network (CDEN) model for probabilistic forecasting using log-normal distributions. The model is trained on historical sediment data, evaluated using multiple random seeds, and assessed using various error metrics.
